---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
#### Load Libraries
```{r, results="markup", message=FALSE, warning=FALSE}
packages <- c("lubridate", "dplyr", "ggplot2", "RcolorBrewer", "mice", "VIM")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")  
}

library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(mice)
library(VIM)
```

## Loading and preprocessing the data
#### Load the Data containing activity data for October and November 2012
```{r, results="markup"}
if (!file.exists('./activity.csv')) {
    unzip('./activity.zip')
}

data <- read.csv("./activity.csv", sep = ",", header = TRUE, fill = TRUE)
```

#### Transform Data - Add day of week
```{r}
data <- data %>% mutate(monthday = format(as.Date(data$date, "%Y-%m-%d"), format = "%m-%d"),
                        dayofweek = weekdays(as.Date(data$date)))
```
## What is mean total number of steps taken per day?

#### Aggregate data by Date
```{r}
stepsPerDay <- select(data, steps, monthday) %>% 
  filter(steps >= 0) %>% 
  group_by(monthday = as.factor(monthday)) %>% 
  summarize(count = n(), 
            mean = mean(steps, na.rm = TRUE),
            median = median(steps),
            totalSteps = sum(steps, na.rm = TRUE))
```

#### Plot the Histogram of the total number of steps taken each day
###### *Note 10/1 was excluded because it contained no values*
```{r}
ggplot(data = stepsPerDay) + 
  geom_histogram(binwidth = 750, color = "white", aes(totalSteps, fill=totalSteps)) +
  labs(x="Total Number of Steps Taken Each Day", y="Frequency",
       title="Histogram of the Total Number of Steps Taken Each Day") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  theme(axis.title = element_text(face="bold")) +
  scale_fill_brewer(palette = "Dark2")
```
###### Calculate the mean and median for the total number of steps taken
```{r}
stepsPerDayMean = format(mean(stepsPerDay$totalSteps), digits = 8)
stepsPerDayMedian = format(median(stepsPerDay$totalSteps), digits = 8)
```
####### **Mean:** `r stepsPerDayMean` steps
####### **Median:** `r stepsPerDayMean` steps

## What is the average daily activity pattern?

#### Aggregate data by time interval (Elimiate Oct 1 due to no values)
```{r}
stepsPerInterval <- filter(data, steps >= 0) %>%
  group_by(interval) %>%
  summarise(avgStepsPerInterval = mean(steps, na.rm = TRUE)) %>%
  mutate(hourlyInterval = interval / 100) %>%
  mutate(hour = floor(hourlyInterval), minute = (hourlyInterval - floor(hourlyInterval)) * 100)
```

#### Time series plot of the average number of steps per 5-Min interval
```{r}
ggplot(data = stepsPerInterval, aes(x = hourlyInterval, y = avgStepsPerInterval)) + 
  geom_line() +
  labs(x="Steps By 5-Min Time Interval (hourly scale)", y="Total Number of Steps",
       title="Average Number of Steps Taken of the 5-Minute Interval") + 
  scale_x_continuous(breaks = scales::pretty_breaks(12)) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  theme(axis.title = element_text(face="bold")) +
  scale_fill_brewer(palette = "Dark2")
```

## Imputing missing values
#### Quickly analize the data to see what the values look like
```{r}
summary(data)
```

#### Give the percentage for each column missing
```{r}
pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(data,2,pMiss)
```

#### Using mice for looking at missing data pattern
```{r, results="markup", message=FALSE, warning=FALSE}
md.pattern(data)
```
##### * The results show the following:
+ 15264 are complete
+ 2304 values are missing from **steps** measurement

##### Give a visual representation using VIM
##### This demonstrates that all missing values belong to **steps** and at the beginning of the data set

```{r}
aggr_plot <- aggr(data, col=c('navyblue','red'),
                  numbers=TRUE, sortVars=TRUE, labels=names(data),
                  cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

marginplot(data[c(1,2)])
```

##### Now that we understand missing values are at the top and scattered for steps
##### We can assume the person did not start until Oct 2.  Oct 2nd was a startup day and the
##### other missing values are days not used.

##### Impute the missing values
```{r results="hide"}
workerImputedData <- mice(data,m=5, maxit=5, meth='pmm',seed=500)
```
```{r}
head(workerImputedData$imp$steps, 30)
imputedData <- complete(workerImputedData, 1)
```
#### Aggregate the imputed data by time interval
```{r}
stepsPerDayImputed <- select(imputedData, steps, monthday) %>% 
  filter(steps >= 0) %>% 
  group_by(monthday = as.factor(monthday)) %>% 
  summarize(count = n(), 
            mean = mean(steps, na.rm = TRUE),
            median = median(steps),
            totalSteps = sum(steps, na.rm = TRUE))
```
#### Time series plot of the average number of steps per 5-Min interval (imputed data)
```{r}
ggplot(data = stepsPerDayImputed) + 
  geom_histogram(binwidth = 750, color = "white", aes(totalSteps, fill=totalSteps)) +
  labs(x="Total Number of Steps Taken Each Day", y="Frequency",
       title="Histogram of the Total Number of Steps Taken Each Day",
       subtitle="with Imputed Values") +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        plot.subtitle = element_text(hjust = 0.5, face = "bold")) +
  theme(axis.title = element_text(face="bold")) +
  scale_fill_brewer(palette = "Dark2")
```

###### Calculate the mean and median for the total number of steps taken on Imputed data
```{r}
stepsPerDayMeanImputed = format(mean(stepsPerDayImputed$totalSteps), digits = 8)
stepsPerDayMedianImputed = format(median(stepsPerDayImputed$totalSteps), digits = 8)
```
####### **Imputed Mean:** `r stepsPerDayMeanImputed` steps
####### **Imputed Median:** `r stepsPerDayMedianImputed` steps

## Are there differences in activity patterns between weekdays and weekends?

##### Transform data by adding the day type for both data and imputed data
```{r}
data$daytype <- ifelse(as.POSIXlt(data$date)$wday %in% c(0,6), 'weekend','weekday')
imputedData$daytype <- ifelse(as.POSIXlt(imputedData$date)$wday %in% c(0,6), 'weekend','weekday')
```

##### Aggregate data by interval and daytype and plot showing weekday verses weekend
```{r}
avgByDaytype <- data %>%
  filter(steps >= 0) %>%
  group_by(interval, daytype) %>%
  summarise(avgStepsPerInterval = mean(steps, na.rm = TRUE)) %>%
  mutate(hourlyInterval = interval / 100) %>%
  mutate(hour = floor(hourlyInterval), minute = (hourlyInterval - floor(hourlyInterval)) * 100)

ggplot(data = avgByDaytype, aes(hourlyInterval, avgStepsPerInterval)) + 
  geom_line() +
  facet_grid(daytype ~ .) +
  scale_x_continuous(breaks = scales::pretty_breaks(12)) +
  labs(x="5-Minute Time Interval (24 hour scale)", y="Avg Number of Steps",
       title="Time Series of Steps Taken",
       subtitle = "Weekday verses Weekend") +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        plot.subtitle = element_text(hjust = 0.5, face = "bold")) +
  theme(axis.title = element_text(face="bold")) +
  scale_fill_brewer(palette = "Dark2")

avgByDaytypeImputed <- imputedData %>%
  filter(steps >= 0) %>%
  group_by(interval, daytype) %>%
  summarise(avgStepsPerInterval = mean(steps, na.rm = TRUE)) %>%
  mutate(hourlyInterval = interval / 100) %>%
  mutate(hour = floor(hourlyInterval), minute = (hourlyInterval - floor(hourlyInterval)) * 100)

ggplot(data = avgByDaytypeImputed, aes(hourlyInterval, avgStepsPerInterval)) + 
  geom_line() +
  facet_grid(daytype ~ .) +
  scale_x_continuous(breaks = scales::pretty_breaks(12)) +
  labs(x="5-Minute Time Interval (24 hour scale)", y="Avg Number of Steps",
       title="Time Series of Steps Taken (Imputed values)",
       subtitle = "Weekday verses Weekend") +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        plot.subtitle = element_text(hjust = 0.5, face = "bold")) +
  theme(axis.title = element_text(face="bold")) +
  scale_fill_brewer(palette = "Dark2")
```
